---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 4,
  dpi = 150
)
```

# spiro

The goal of `spiro` is to enable a fast workflow with raw files from metabolic
carts/spiroergometry systems. It provides simple tools for data
import, processing, summary and visualisation.

## Installation

Install the current development version from GitHub:

``` {r installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("smnnlt/spiro")
```

## Usage

* Use `spiro()` for one-step data import and processing from raw metabolic cart
  files. 
* Use `spiro_summary()` and `spiro_glance()` for getting a stepwise or
  overall summary of the imported data. 
* Use `spiro_plot_*()` as a shortcut for visualising the imported data.
  
#### Metabolic Carts

The following metabolic carts are currently supported by `spiro`:

* COSMED
* ZAN

## Example

```{r example, warning=FALSE}
library(spiro)

# get data for example
file <- spiro_example("zan_gxt")

gxt_data <- spiro(file)
spiro_summary(gxt_data)

spiro_plot_VO2(gxt_data)
```
