---
title: "Summarising & Plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising & Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.width = 6,
  fig.asp = 0.618,
  fig.align = "center",
  dpi = 150
)
```

Different measuring devices and analysis software lead to opaque results in measuring gas exchange parameters. To make exercise science more transparent and reproducible, the `spiro` package offers a standardised workflow for data from metabolic carts.

This vignette provides you information on how to summarise and plot data previously imported and processed by `spiro()`.

```{r setup}
library(spiro)

# import and process example data
file <- spiro_example("zan_gxt")
gxt_data <- spiro(file)
gxt_data

```

## Stepwise summary with `spiro_summary()`

To analyse gas exchange data, usually mean parameters for each performed load step are warranted. To ensure the presence of a metabolic steady state, the end of each step is used for calculations.

```{r spiro_summary}
spiro_summary(gxt_data, interval = 120)
```

The length of the computational interval in seconds can be modified by setting the `interval` argument. If the interval exceed the length of any step, it will be shortened in these cases and you receive a note.

## Overall summary with `spiro_glance()`

For some types of exercise tests it may be preferable to get an overall instead of a stepwise summary of parameters. `spiro_glance` will usually show the maximal values within the test after applying a smoothing by the use of a rolling average.

```{r spiro_glance}
spiro_glance(gxt_data, interval = 30)
```

In this case, the `interval` argument will set the length over which the rolling average will be computed.

For constant load test `spiro_glance()` will behave different, as not the maximal values, but the overall average is useful. Therefore it will calculate the mean parameter values for all steps of the constant load.

## Plotting the data

To plot exercise test data retrieved by `spiro()` it is recommend to use specific visualisation packages such as `ggplot2`, which provide a variety of possible applications.
However, to provide a quick view on the compiled data (e.g. for checking if the import operated correctly) `spiro` offers function for data visualisation.

### Plotting oxygen uptake data

In many cases, the (relative) oxygen uptake will be the parameter of most interest when analysing spiroergometric data. `spiro_plot_VO2` visualises the interpolated oxygen uptake data, supplemented by it's rolling average and the underlying protocol load.

The length of the computational interval for the rolling average can be changed by setting the `smooth` argument. To allow a further check for correct importing, setting the `title` argument to `TRUE` will create a plot title with the participant's name and the test type.

```{r spiro_plot_VO2, warning=FALSE}
spiro_plot_VO2(gxt_data, 
               smooth = 15,
               title = FALSE
               )
```

### Plotting heart rate data

To verify the presence of correct heart rate data, `spiro_plot_HR` provides an easy option of visualising the data together wwith the protocol load data.

To allow a further check for correct importing, setting the `title` argument to `TRUE` will create a plot title with the participant's name and the test type.

```{r spiro_plot_HR, warning=FALSE}
#spiro_plot_HR(gxt_data, 
#               title = FALSE
#               )
```
